---
title: "Prediction on Wearable Device dataset"
author: "Eshu English"
date: "9 April 2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, results = 'asis')
```

## Instructions
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.

In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website [ here](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).  
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

## Data
The training data for this project are available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)  
The test data are available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

## Review criteria
  1. Your submission for the Peer Review portion should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
  2. Apply your machine learning algorithm to the 20 test cases available in the test data above and submit your predictions in appropriate format to the Course Project Prediction Quiz for automated grading.

## Analysis
### Environment setup
```{r, message=FALSE}
library(caret)
library(ggplot2)
library(dplyr)
if(!file.exists('train.csv')){
        download.file(url = 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv',
                      destfile = 'train.csv',method = 'curl', quiet = TRUE)
}
if(!file.exists('test.csv')){
        download.file(url = 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv',
                      destfile = 'test.csv',method = 'curl', quiet = TRUE)
}

training <-  read.csv('train.csv')
testing <- read.csv('test.csv')
```

### Preprocessing
1. Examining the structure of the data and removing variables unrelated to exercise (column number and time stamps)
```{r}
str(training)
training <- select(training,6:ncol(training))
```
2. Split the data into 70% training and 30% testing set and setting seed for reproducibility

```{r}
set.seed(1111)
inTrain <- createDataPartition(training$classe, p = 0.7, list = FALSE)
trainSet<- training[inTrain,]
testSet <- training[-inTrain,]
```
3. Removing features having little value
```{r}
nsv <- nearZeroVar(trainSet, saveMetrics = TRUE)
impFeat <- row.names(nsv[nsv$nzv ==FALSE,])
trainSet<- trainSet[,impFeat]
```
4.Removing the features with all NAs
```{r}
trainSet<- trainSet[,colSums(is.na(trainSet))==0]
ncol(trainSet)
```
There are still huge number of features i.e. 54

### Model training
```{r, include=FALSE}
rm(training)
```
1. Set up 5-fold cross validation for training
```{r}
set.seed(111)
crossVal<- trainControl(method = 'cv', number = 5)
```

2. Fitting the model using Random Forest.
```{r}
set.seed(111)
modelRF <- train(classe~., data = trainSet,method = 'rf', trcontrol = crossVal)
modelRF$finalModel
```

- Prediction on the test set and check the confusion matrix and accuracy.
```{r}
predRf <- predict(modelRF, newdata = testSet)
confusionMatrix(predRf, testSet$classe)$table
confusionMatrix(predRf, testSet$classe)$overall[1]
```
The Accuracy is about 99% using Random Forest classifier

3. Fitting the model using Gradient Boosting.
```{r}
modelGBM <- train(classe ~., data = trainSet, method = 'gbm', trControl = crossVal, verbose = F)
modelGBM$finalModel
```
- Prediction on the test set and check the confusion matrix and accuracy.
```{r}
predGBM <- predict(modelGBM, newdata = testSet)
confusionMatrix(predGBM, testSet$classe)$table
confusionMatrix(predGBM, testSet$classe)$overall[1]
```
The Accuracy is about 98% using Gradient Boosting as a classifier.(A little lower than Random Forest)

## Now Predicting on th validation set, loaded by the name of testing
Prediction done using Random Forest cos it performed better on the test set above
```{r}
predVal <- predict(modelRF, newdata = testing)
predVal
```

